---
title: "Data on US Federal Balance of Payments, 2018"
author: "Joseph Nathan Cohen"
date: "12/9/2020"
output: 
  html_document:
    toc: true
  
---

```{r setup, include=FALSE}

# Clear Memory
rm(list=ls())
gc()

# Set Working Directory
directory <- "E:/Dropbox/Research/State Balance of Payments"
data.directory <- paste0(directory, "/Data")
setwd(directory)

# Set Random Seed
set.seed(123)

# Turn off Scientific Notation
options(scipen=999)

# Load R packages used in this analysis
library(dplyr)
library(readxl)
library(openxlsx)
library(kableExtra)
library(formattable)
library(scales)
library(maps)
library(data.table)
library(ggsn) 
library(ggrepel)
library(usmap)
library(ggplot2)
library(lmtest)
library(sandwich)
library(stargazer)
library(psycho)
library(ggmap)
library(maps)
library(mapdata)
library(tidyr)
library(dplyr)

```

The *US Federal Balance of Payments* data set is a compilation of state-level public finance, demographic, geographic, and other data, covering the year 2018.  This document outlines the variables contained in this set, and the sources from which they were drawn. 

Download the data set in comma-separated values format here.

The scripts and raw data used to consolidate this data are available for download on this project's archive on the [Open Science Framework](https://osf.io/8cqtg/)^[Cohen, Joseph N. 2020. “State Balance of Payment Data, 2018.” Retrieved <osf.io/eh2d9>.]

# Data & Sources

This section describes the data set and the sources used to build it.

## Federal Balance-of-Payment Data

The set is centered on a state-disaggregated federal revenue and expenditure set maintained by the [New York Comptroller's Office](https://www.osc.state.ny.us/files/reports/budget/pdf/federal-budget-fiscal-year-2018.pdf).^[New York State Comptroller’s Office (2020) New York's Balance of Payments in the Federal Budget, January <https://www.osc.state.ny.us/files/reports/budget/pdf/federal-budget-fiscal-year-2018.pdf>.] Its associated variables are described in the table below:

```{r NY state data, include = F}
# NYS Balance of Payment Data
# Data is distributed in Excel workbook spread across multiple sheets
# Below, I tidy them into a unified data table
setwd(data.directory)

temp.1 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 2, skip = 2)
names(temp.1) <- paste(c("state", "bop.total", "bop.pc", "DROP"))
temp.1 <- temp.1[,c(1:3)]

temp.2 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 4, skip = 2)
names(temp.2) <- paste(c("state", "tax.paid.total", "tax.paid.pc", "DROP"))
temp.2 <- temp.2[,c(1:3)]

temp.3 <- merge(temp.1, temp.2, by = "state")

temp.4 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 5, skip = 3)
names(temp.4) <- paste(c("state", "tax.paid.pinc", "tax.paid.pinc.pc", "DROP", "DROP",
                         "tax.paid.socins", "tax.paid.socins.pc", "DROP","DROP",
                         "tax.paid.corp", "tax.paid.corp.pc", "DROP","DROP",
                         "tax.paid.excise", "tax.paid.excise.pc", "DROP","DROP",
                         "tax.paid.estate", "tax.paid.estate.pc", "DROP"))
temp.4 <- temp.4[,c(1:3,6,7,10,11,14,15,18,19)]

temp.5 <- merge(temp.3, temp.4, by = "state")

temp.6 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 6, skip = 2)
names(temp.6) <- paste(c("state", "exp.total", "exp.pc", "DROP"))
temp.6 <- temp.6[,c(1:3)]

temp.7 <- merge(temp.5, temp.6, by = "state")

temp.8 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 7, skip = 2)
names(temp.8) <- paste(c("state", "exp.dirpmt.total", "exp.dirpmt.pc", "DROP"))
temp.8 <- temp.8[,c(1:3)]

temp.9 <- merge(temp.7, temp.8, by = "state")

temp.10 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 8, skip = 3)
names(temp.10) <- paste(c("state", "exp.dirpmt.socsec.tot", "exp.dirpmt.socsec.pc", "DROP", "DROP",
                          "exp.dirpmt.medicare.tot", "exp.dirpmt.medicare.pc", "DROP", "DROP",
                          "exp.dirpmt.vetben.tot", "exp.dirpmt.vetben.pc", "DROP", "DROP",
                          "exp.dirpmt.fedempl.tot", "exp.dirpmt.fedempl.pc", "DROP", "DROP",
                          "exp.dirpmt.snap.tot", "exp.dirpmt.snap.pc", "DROP", "DROP",
                          "exp.dirpmt.eitc.tot", "exp.dirpmt.eitc.pc", "DROP", "DROP",
                          "exp.dirpmt.ssi.tot", "exp.dirpmt.ssi.pc", "DROP"))
temp.11 <- temp.10[,c(1:3,6,7,10,11,14,15,18,19,22,23,26,27)]
temp.12 <- merge(temp.9, temp.11, by = "state")

temp.13 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 9, skip = 2)
names(temp.13) <- paste(c("state", "exp.grants.total", "exp.grants.pc", "DROP"))
temp.13 <- temp.13[,c(1:3)]
temp.14 <- merge(temp.12, temp.13, by = "state")

temp.15 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 10, skip = 3)
names(temp.15) <- paste(c("state", "exp.grants.medicaid.tot", "exp.grants.medicaid.pc", "DROP", "DROP",
                          "exp.grants.hiways.tot", "exp.grants.hiways.pc", "DROP", "DROP",
                          "exp.grants.housing.tot", "exp.grants.housing.pc", "DROP", "DROP",
                          "exp.grants.chldnut.tot", "exp.grants.chldnut.pc", "DROP", "DROP",
                          "exp.grants.tanf.tot", "exp.grants.tanf.pc", "DROP", "DROP",
                          "exp.grants.esea.tot", "exp.grants.esea.pc", "DROP", "DROP",
                          "exp.grants.sped.tot", "exp.grants.sped.pc", "DROP", "DROP",
                          "exp.grants.chip.tot", "exp.grants.chip.pc", "DROP", "DROP",
                          "exp.grants.transit.tot", "exp.grants.transit.pc", "DROP"))
temp.15 <- temp.15[,c(1:3,6,7,10,11,14,15,18,19,22,23,26,27,30,31,34,35)]
temp.16 <- merge(temp.14, temp.15, by = "state")

temp.17 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 11, skip = 2)
names(temp.17) <- paste(c("state", "exp.proc.total", "exp.proc.pc", "DROP"))
temp.17 <- temp.17[,c(1:3)]
temp.18 <- merge(temp.16, temp.17, by = "state")

temp.19 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 12, skip = 3)
names(temp.19) <- paste(c("state", "exp.proc.mil.tot", "exp.proc.mil.pc", "DROP", "DROP",
                          "exp.proc.nonmil.tot", "exp.proc.nonmil.pc", "DROP"))
temp.19 <- temp.19[,c(1:3,6,7)]
temp.20 <- merge(temp.18, temp.19, by = "state")

temp.21 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 13, skip = 2)
names(temp.21) <- paste(c("state", "exp.wages.total", "exp.wages.pc", "DROP"))
temp.21 <- temp.21[,c(1:3)]
temp.22 <- merge(temp.20, temp.21, by = "state")

temp.23 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 14, skip = 3)
names(temp.23) <- paste(c("state", "exp.wages.mil.tot", "exp.wages.mil.pc", "DROP", "DROP",
                          "exp.wages.civ.tot", "exp.wages.civ.pc", "DROP"))
temp.23 <- temp.23[,c(1:3,6,7)]
temp.24 <- merge(temp.22, temp.23, by = "state")

temp.25 <- read_xlsx("NYS Federal Payments Data.xlsx", sheet = 15, skip = 2)
temp.25 <- temp.25[,c(1,3)]
names(temp.25) <- paste(c("state", "population"))
temp.26 <- merge(temp.24, temp.25, by = "state")

bop.data <- temp.26
rm(list=ls(pattern="temp"))

var.descr <- c(
  "State's federal balance-of-payments",
  "Per capita balance-of-payments",
  "Taxes paid, total",
  "Taxes paid, per capita",
  "Taxes paid: Personal, total",
  "Taxes paid: Personal, per capita",
  "Taxes paid: Social Insurance, per capita",
  "Taxes paid: Social Insurance, total",
  "Taxes paid: Corporate, total",
  "Taxes paid, Corporate, per capita",
  "Taxes paid: Excise, total",
  "Taxes paid, Excise, per capita",
  "Taxes paid, Estate, total",
  "Taxes paid: Estate, per capita",
  "Expenditures, total",
  "Expenditures, per capita",
  "Expenditures: Direct payments, total",
  "Expenditures: Direct payments, per capita",
  "Expenditures: Direct payments: Social Security, total",
  "Expenditures: Direct payments: Social Security, per capita",
  "Expenditures: Direct payments: Medicare, total",
  "Expenditures: Direct payments: Medicare, per capita",
  "Expenditures: Direct payments: Veterans' Benefits, total",
  "Expenditures: Direct payments: veterans' benefits, per capita",
  "Expenditures: Direct payments: federal employment, total",
  "Expenditures: Direct payments: federal employment, per capita",
  "Expenditures: Direct payments: SNAP, total",
  "Expenditures: Direct payments: SNAP, per capita",
  "Expenditures: Direct payments: EITC, total",
  "Expenditures: Direct payments: EITC, per capita",
  "Expenditures: Direct payments: SSI, total",
  "Expenditures: Direct payments: SSIR, per capita",
  "Expenditures: Grants, total",
  "Expenditures: Grants, per capita", 
  "Expenditures: Grants: Medicaid, total",
  "Expenditures: Grants: Medicaid, per capita",
  "Expenditures: Grants: Highways, total",
  "Expenditures: Grants: Highways, per capita", 
  "Expenditures: Grants: Housing, total",
  "Expenditures: Grants: Housing, per capita", 
  "Expenditures: Grants: Child Nutrition, total",
  "Expenditures: Grants: Child Nutrition, per capita", 
  "Expenditures: Grants: TANF, total",
  "Expenditures: Grants: TANF, per capita", 
  "Expenditures: Grants: ESEA, total",
  "Expenditures: Grants: ESEA, per capita", 
  "Expenditures: Grants: Special Ed, total",
  "Expenditures: Grants, Special Ed, per capita", 
  "Expenditures: Grants: CHIP, total",
  "Expenditures: Grants: CHIP, per capita", 
  "Expenditures: Grants: Transit, total",
  "Expenditures: Grants: Transit, per capita", 
  "Expenditures: Procurement, total",
  "Expenditures: Procurement, per capita", 
  "Expenditures: Procurement: Military, total",
  "Expenditures: Procurement: Military, per capita",
  "Expenditures: Procurement: Non-Military, total",
  "Expenditures: Procurement: non-military, per capita",
  "Expenditures: Wages, Total",
  "Expenditures: Wages, per capita",
  "Expenditures: Wages: military, total",
  "Expenditures; Wages: military, per capita",
  "Expenditures: Wages, civilian, total",
  "Expenditures: Wages, civilian, per capita")

Names <- names(bop.data)[-c(1,66)]
tab.bop.dat <- 
  cbind(Names,var.descr)
names(tab.bop.dat) <- paste(c("Variable", "Explanation"))
  
rm(Names, var.descr)
  
```

```{r NY state BOP data variables, include = T, echo = F, results = "asis"}

kable(tab.bop.dat, booktabs = T, longtable = T, caption = "Federal Transfers Variables") %>%
  kable_styling()

```

## State Population & Density Data

We use a Census distribution for state-level population, population growth, and density data.^[U.S. Census Bureau “Table 14. State Population - Rank, Percent Change, and Population Density” Data set downloaded November 1, 2020 from https://www2.census.gov/library/publications/2011/compendia/statab/131ed/tables/12s0014.xls>]  To examine the role that large city populations, we used Census data to score the percentage of a state's population residing in a Combined Statistical Area with at least 2 million people.  These variables are all described below:

```{r Census population and density data set up, include = F}

setwd(data.directory)

census.pop <- read_xls("Census Pop and Density.xls", skip = 4)
census.pop <- census.pop[-c(1:2),c(1,18,24)]
names(census.pop) <- paste(c("state", "pop.growth", "pop.density"))
bop.data <- merge(bop.data, census.pop, by = "state")

```

```{r setup megacity population data, include = F}
setwd(data.directory)

temp.1 <- read.csv("Census CSA Data.csv")

temp.2 <- subset(temp.1, LSAD == "Combined Statistical Area" & 
                   CENSUS2010POP > 2000000)[c(1,2,5,7)]
tab.csa.pops <- arrange(temp.2[c(3,4)], -temp.2[4])


temp.csa <- temp.2$CSA
temp.3 <- subset(temp.1, LSAD != "Combined Statistical")

for (i in temp.csa){
  temp.A <- subset(temp.3, CSA == i)
  temp.cbsas <- unique(temp.A$CBSA)
  temp.B <- subset(temp.1, LSAD == "County or equivalent" & CBSA %in% temp.cbsas)[c(5,7)]
  temp.vec <- as.character(temp.B$NAME)
  temp.C <- unlist(strsplit(temp.vec, ", "))
  temp.B$COUNTY <- temp.C[seq(1, length(temp.C), 2)]
  temp.B$STATE <- temp.C[seq(2, length(temp.C), 2)]
  temp.vecB <- unique(temp.B$STATE)
  for (j in temp.vecB){
    temp.C <- subset(temp.B, STATE == j)
    assign(paste0("temD.",i,".",j), sum(temp.C$CENSUS2010POP))
  }
}

rm(list=ls(pattern="temp"))

for (i in state.abb){
  temp.1 <- grep(i,names(.GlobalEnv),value=TRUE)
  temp.dat <- 0
  
  
  for (j in 1:length(temp.1)){
    temp.dat <- ifelse(is.na(temp.1[j]), temp.dat, temp.dat + get(temp.1[j]))
  }
  assign(paste0("temp.", i), temp.dat)
}

for (i in state.abb){
  temp <- get(paste0("temp.", i))
  temp <- ifelse(is.logical(temp), 0, temp)
  assign(paste0("temp.", i), temp)
}

temp <- mget(paste0("temp.", state.abb)) 

temp.data <- data.frame(state.abb = state.abb, 
                        state = state.name,
                        metro.pop = array(as.numeric(unlist(temp)), dim=c(50,1)))

bop.data <- merge(bop.data, temp.data, by = 'state')
bop.data$metro.pop.pct <- (bop.data$metro.pop / bop.data$population) * 100

rm(list=ls(pattern = "temD"))

# What % in donor state megacities in 2010?
(22255491 + 17877006 + 9840929 + 8923942 + 7893376 + 3706278)/309300000

```

```{r variable table - census data, echo = F, results = "asis"}

tab <- data.frame(
  Variable = c("population", "pop.growth", "pop.density", "metro.pop", "metro.pop.pct"),
  Explanation = c("Total population", "Population Growth (%)", "State Population Density", 
    "Metropolitan Population", "Metro Population (%)"))

kable(tab, booktabs = T, longtable = T, caption = "Federal Transfers Variables") %>%
  kable_styling()

rm(list=ls(pattern="tab"))

```
## State GDP Data

State GDP data was drawn from the [Bureau of Economic Analysis](https://apps.bea.gov/itable/iTable.cfm?ReqID=70&step=1#reqid=70&step=1&isuri=1).^[United States Bureau of Economic Analysis (2020) “Regional Data: GDP and Personal Income” Online dataset. Accessed November 11, 2020.  <https://apps.bea.gov/itable/iTable.cfm?ReqID=70&step=1#reqid=70&step=1&isuri=1>]  The set includes data on overall GDP, and government value-added.  It also breaks down government value-added to federal civilian, state/local civilian, and military.

```{r set up gdp data, include = F}
setwd(data.directory)
temp.1 <- read.csv("BEA State GDP & Govt VA.csv")
temp.1 <- temp.1[-c(1:9),-c(1,3)]
temp.gdp <- subset(temp.1, X.2 == "All industry total")[c(1,3)]
names(temp.gdp) <- paste(c("state", "gdp"))
temp.govVA <- subset(temp.1, X.2 == "  Government and government enterprises")[c(1,3)]
names(temp.govVA) <- paste(c("state", "gov.va"))
temp.govVA.fedciv <- subset(temp.1, X.2 == "    Federal civilian")[c(1,3)]
names(temp.govVA.fedciv) <- paste(c("state", "gov.va.fedciv"))
temp.govVA.mil <- subset(temp.1, X.2 == "    Military")[c(1,3)]
names(temp.govVA.mil) <- paste(c("state", "gov.va.mil"))
temp.govVA.sl <- subset(temp.1, X.2 == "    State and local")[c(1,3)]
names(temp.govVA.sl) <- paste(c("state", "gov.va.sl"))

temp.2 <- merge(temp.gdp, temp.govVA, by = "state")
temp.2 <- merge(temp.2, temp.govVA.fedciv, by = "state")
temp.2 <- merge(temp.2, temp.govVA.mil, by = "state")
temp.2 <- merge(temp.2, temp.govVA.sl, by = "state")

bop.data <- merge(bop.data, temp.2, by = "state")

bop.data$gdp.pc <- (bop.data$gdp * 1000) / bop.data$population
bop.data$gdp.pc.K <- bop.data$gdp.pc / 1000
  
rm(list=ls(pattern="temp"))

```

```{r VA table, echo = F, include = T}

tab <- data.frame(Variables = names(bop.data)[72:77],
                  Description = c(
                    "State GDP", "Total Government Value-Added", 
                    "Govt Value;Added: Federal Civilian",
                    "Govt Value-Added: Military",
                    "Govt Value-Added: State & Local",
                    "GDP per capita"
                  ))

kable(tab, booktabs = T, longtable = T, caption = "Economic Value-Added Variables") %>%
  kable_styling()
```

## Census Area

R's base package includes data on state's geographic size and the Census divisions to which states pertain.

```{r census area data setup, include = F}

temp.1 <- data.frame(state.abb, state.area, state.division, state.region)
bop.data <- merge(bop.data, temp.1, by = "state.abb")

```

```{r state abb table, echo = F, include = T}

tab <- data.frame(Variables = names(bop.data)[79:81],
                  Description = c(
                    "State area in mi^2",
                    "Census-defined Division",
                    "Census-defined Region"))

kable(tab, booktabs = T, longtable = T, caption = "Area and Region Variables") %>%
  kable_styling()

```

## Poverty

The set includes two poverty metrics, drown from the [U.S Department of Agriculture's Economic Research Service](https://data.ers.usda.gov/reports.aspx?ID=17826)^[U.S. Department of Agriculture (2020) "Home / Data Products / County-level Data Sets / Poverty" Electronic data set downloaded December 9, 2020 from <https://data.ers.usda.gov/reports.aspx?ID=17826>]

```{r poverty data setup, include = F}
setwd(data.directory)
temp.1 <- read.csv("USDA Poverty Estimates.csv")
temp.1 <- temp.1[c(1,2,5)]
names(temp.1) <- paste(c("state", "pov.pct", "pov.u18.pct"))
bop.data <- merge(bop.data, temp.1, by = "state")

```

```{r table poverty, include = T, echo = F}
tab <- data.frame(Variable = c("pov.pct", "pov.u18.pct"),
                  Explanation = c("Poverty Rate", "Under-18 Poverty Rate"))
kable(tab, booktabs = T, longtable = T, caption = "Poverty Variables") %>%
  kable_styling()
```

## Local Prices

Local price data is drawn from the [U.S. Bureau of Economic Analysis](https://apps.bea.gov/iTable/iTable.cfm?reqid=70&step=1&isuri=1&acrdn=8#reqid=70&step=1&isuri=1).^[U.S. Bureau of Economic Analysis (2020) "Regional Data: SARPP Regional Price Parities by State" Electronic data set downloaded December 9, 2020 from <https://apps.bea.gov/iTable/iTable.cfm?reqid=70&step=1&isuri=1&acrdn=8#reqid=70&step=1&isuri=1>]

```{r regional prices data setup, include = F}
setwd(data.directory)
temp.1 <- read.csv("BEA COL Data.csv")
temp.1 <- temp.1[2:3]
names(temp.1) <- paste(c("state", "costliving"))
bop.data <- merge(bop.data, temp.1, by = "state")

```

```{r table prices, include = T, echo = F}
tab <- data.frame(Variable = c("pov.pct", "pov.u18.pct"),
                  Explanation = c("Poverty Rate", "Under-18 Poverty Rate"))
kable(data.frame(Variable = "costliving", Explanations = "State cost of living"), 
      booktabs = T, longtable = T, caption = "Poverty Variables") %>%
  kable_styling()
```

# Citing this Set

Please cite this set as:

Cohen, Joseph N. 2020. “State Balance of Payment Data, 2018.” Data set. <osf.io/eh2d9>

```{r write data set to drive, include = F}

# Last-Minute rescales
bop.data$state.area <- bop.data$state.area / 1000

#Writing data set
setwd(directory)
write.xlsx(bop.data, file = "US State Federal Payments 2018.xlsx")
write.csv(bop.data, file = "US State Federal Payments 2018.csv")


```
